# (PART) Introduction {-}

About Time Series
====================

What is time series?
--------------------

1. Observations (obs.) could be 

* **univariate** or **multiivariate**
* **discrete** (count) or **continuous** (measurement)

2. Set of obs. recorded *sequentially* (has a natural order)

* **time**
* **spatial ordering**, eg. $1^{st}$ plaint in a row, $2^{nd}$ plaint in a row
* **depth**, eg. soil (土壤) PH: 1m down, 2m down

3. Recording *interval* could be

* **regular**, eg. hourly, daily, monthly
* **irregular**, eg. earthquake, observe stock price at transaction times
  * take average or median to change it into regular

Plot of time series data
--------------------

plot *data vs time* to look:

* **trend**: upward or downward (整體趨勢)
* **repetitive pattern** is unknown or unknown:
  * **periodicity**: in regular pattern, eg. 1.3 years
  * **seasonality**: known period, eg. 1 year, 12 month for mothly data
* **heteroskadesticity $\sigma^2_t$**: changing variance
* **dependence**: positive or negative
  * successive obs. ($x_t, x_{t+1}$) are similar or dissimilar
```{r echo=FALSE}
par(mfrow=c(1,2))
plot(1:5, c(0,1,2,1,0), type = 'b', ylim = c(0,3), xlab = '', ylab = '', main = 'positive dependence')
plot(1:5, c(0,1,0.5,2,1.5), type = 'b', ylim = c(0,3), xlab = '', ylab = '', main = 'negative dependence')
```
* missing, outliers

Why use time series
--------------------

* difference between Regression (reg.) and Time Series (TS)
  * **Reg.**: need (X, Y)  
    X之間獨立 (independent, indep.), X和Y之間不獨立 (dependent, dep.)
  * **TS**: only need X (times)  
    X之間 dep.

1. Often in time series data:

* trend 不好使用 global regression function 來 fitted
  * 如下圖，就需要2個 reg.
```{r echo=FALSE}
plot(1:3, c(1,2,1), type = 'l', ylim = c(1,3), xlab = '', ylab = '')
```
* display **time-varying** local behavior
* often have seasonality

2. *Ignore dependence* often lead to:

* **inefficient** estimates of reg. parameters
* poor prediction
* standard errors (std.): unrealistically small
  * confidence interval (CI.): unrealistically narrow
  * hypothesis testing: improper inferences

Objective of time series analysis
--------------------

1. provide an *interpretable* model of data

* often involve **multivariate** series
* test if the variable is **significance (sig.)** influence

2. *predict* future values of series

* predictive model often do not try to explain

Modeling strategy
--------------------

$$\hat{y}_{t+1}=trend+dependence$$
$$trend \leftarrow reg., \:dependence \leftarrow TS$$

1. *difficulties*

* random variables (r.v.) **not identically** distributed
* r.v. **not independent**
* **different means** due to trend or seasonality
* **different variance**

2. try to *make it easier*

* to focus on **modeling dependence**
  * eliminate **trend** and **seasonality**
  * eliminate **changing variance**
* model remainder as **dependent** but **identically** distributed