---
title: "Time Series Example"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
    theme: united
    css: style for united.css
date: '`r Sys.Date()`'
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center',
                      fig.height = 6, fig.width = 12,
                      results = "hold")
```

```{r}
library(fracdiff)
library(astsa)
```

## Multivariate time series data
### <mark>Example 5.8 (p.) </mark>
* `soi.d`: $y_t$
* `soi`: $X_t$
* `time(soi)`: $t$

$$y_t=X_t-\beta_0-\beta_1t$$

```{r results='hide'}
soi.d = resid(lm(soi~time(soi), na.action=NULL)) # detrended SOI
acf2(soi.d)
```

$\uparrow$ PACF cut-off after lag 1 $\Rightarrow$ AR(1)

$\downarrow$ Fit AR(1) model to $y_t$
```{r}
fit = arima(soi.d, order=c(1,0,0))
```

$\downarrow$ $\hat\phi_1$
```{r}
ar1 = as.numeric(coef(fit)[1]) # = 0.5875
```

* $\color{red}{\tilde y_t}$ `soi.pw` $=y_t-\hat\phi_1y_{t-1}$
* $\color{red}{\tilde Z_t}$ `rec.fil` $=Z_t-\hat\phi_1Z_{t-1}$ 
```{r}
soi.pw = resid(fit)
rec.fil = filter(rec, filter=c(1, -ar1), sides=1)
```

`ccf(x, y, ...)` $=corr(x_{t+h},y_t)$

```{r}
ccf(soi.pw, rec.fil, ylab="CCF", na.action=na.omit, panel.first=grid())
```

$\uparrow$ $X_{t-5}$ and $y_t$ are highly correlated. ($Corr(\tilde Y_{t-5}, \tilde Z_t)$) $\Rightarrow$ Use $X_{t-5}$ to predict $y_t$

### <mark>Example 5.9 (p.) </mark>
```{r}
soi.d = resid(lm(soi~time(soi), na.action=NULL))
```

$$fish=\Big[\stackrel{Z_t}{\small\sim}\;\stackrel{Z_{t-1}}{\small\sim}\;\stackrel{y_{t-5}}{\small\sim}\Big]$$

* `rec`: $Z_t$
* `lag(rec,-1)`: $Z_{t-1}$
* `lag(soi.d,-5)`: $y_{t-5}$, choose this because of the CCF in Example 5.8
```{r}
fish = ts.intersect(rec, RL1=lag(rec,-1), SL5=lag(soi.d,-5))
```

`u`: $Z_t=\beta_1 Z_{t-1}+\beta_2 y_{t-5}$
```{r}
(u = lm(fish[,1]~fish[,2:3], na.action=NULL))
```

`resid(u)`: $\mu_t=Z_t-\beta_1 Z_{t-1}-\beta_2 y_{t-5}$
```{r results='hide'}
acf2(resid(u)) # suggests ar1
```

$\uparrow$ PACF cut-off after lag 1 $\Rightarrow$ AR(1)

$\downarrow$ Fit AR(1) to $\mu_t$, $\therefore\mu_t=\phi_1\mu_{t-1}+W_t$
```{r results='hide'}
(arx = sarima(fish[,1], 1, 0, 0, xreg=fish[,2:3])) # final model
```

$\uparrow$ The final model is not good but acceptable. 

```{r}
pred = rec + resid(arx$fit) # 1-step-ahead predictions
ts.plot(pred, rec, col=c('gray90',1), lwd=c(7,1))
legend("topleft", legend=c("Pred.", "True"),
       col=c('gray90',1), lwd=c(7,1))
```

