---
title: "Time Series Example"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
    theme: united
    css: style for united.css
date: '`r Sys.Date()`'
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center',
                      fig.height = 6, fig.width = 12,
                      results = "hold")
library(astsa)
```

### <mark>Example 4.14 (p.) No Taper</mark>
$\downarrow$ Smooth data
```{r}
soi.ave = mvspec(soi, kernel('daniell',4))
abline(v=c(.25,1,2,3), lty=2)
soi.ave$bandwidth # = 0.225
```

* bandwidth: the range of the curve
  * $=\sigma^2$ in Gaussian kernel

$\downarrow I(\omega_j)$
```{r}
df = soi.ave$df # df = 16.9875 (returned values)
U = qchisq(.025, df) # U = 7.555916
L = qchisq(.975, df) # L = 30.17425
```
```{r}
soi.ave$spec[10] # 0.0495202
soi.ave$spec[40] # 0.1190800
```

$\downarrow$ intervals

the CI is smaller than CI in example 4.13
```{r}
df*soi.ave$spec[10]/L # 0.0278789
df*soi.ave$spec[10]/U # 0.1113333
```
```{r}
df*soi.ave$spec[40]/L # 0.0670396
df*soi.ave$spec[40]/U # 0.2677201
```

### <mark>Example 4.16 (p.) Taper=0.1</mark>
```{r}
kernel("modified.daniell", c(3,3))
plot(kernel("modified.daniell", c(3,3)))
```
```{r}
k = kernel("modified.daniell", c(3,3))
soi.smo = mvspec(soi, kernel=k, taper=.1, log="no")
abline(v=c(.25,1), lty=2)
```
```{r}
df = soi.smo$df # df = 17.42618
soi.smo$bandwidth # B = 0.2308103
```
```{r}
soi.smo = mvspec(soi, taper=.1, spans=c(7,7))
```

### <mark>Example 4.18 (p.)</mark>
```{r}
spaic = spec.ar(soi, log="no") # min AIC spec
abline(v=frequency(soi)*1/52, lty=3) # El Nino peak
```

$\downarrow$ Fit AR(1), AR(2), ..., AR(30)

AR(15) is selected. 
```{r}
(soi.ar = ar(soi, order.max=30)) # estimates and AICs

plot(1:30, soi.ar$aic[-1], type="o") # plot AICs
```

$\downarrow$ the minimum AIC/BIC specifies the best model
```{r}
n = length(soi)
AIC = rep(0, 30) -> AICc -> BIC
for (k in 1:30){
sigma2 = ar(soi, order=k, aic=FALSE)$var.pred
BIC[k] = log(sigma2) + (k*log(n)/n)
AICc[k] = log(sigma2) + ((n+k)/(n-k-2))
AIC[k] = log(sigma2) + ((n+2*k)/n) }
IC = cbind(AIC, BIC+1)
ts.plot(IC, type="o", xlab="p -> AR(p)", ylab="AIC / BIC")
```

**Remember residual diagnostic**

### <mark>Example 4.22 (p.)</mark>
$\downarrow$ compare raw data, first diff. and kernel weights
```{r}
par(mfrow=c(3,1), mar=c(3,3,1,1), mgp=c(1.6,.6,0))
plot(soi) # plot data
plot(diff(soi)) # plot first difference
k = kernel("modified.daniell", 6) # filter weights
plot(soif <- kernapply(soi, k)) # plot 12 month filter
```

$\downarrow$ spectral analysis
```{r}
spectrum(soif, spans=9, log="no")
abline(v=12/52, lty="dashed")
```

$\downarrow$ frequency responses

```{r}
par(mfrow=c(2,1), mar=c(3,3,1,1), mgp=c(1.6,.6,0))
w = seq(0, .5, by=.01)
FRdiff = abs(1-exp(2i*pi*w))^2
plot(w, FRdiff, type='l', xlab='frequency')
u = cos(2*pi*w)+cos(4*pi*w)+cos(6*pi*w)+cos(8*pi*w)+cos(10*pi*w)
FRma = ((1 + cos(12*pi*w) + 2*u)/12)^2
plot(w, FRma, type='l', xlab='frequency')
```

$\rightarrow$ UP: FRdiff
$$
\Psi(B)=1-B\\
PTF=\big|\Psi(e^{i\omega})\big|^2=\big|1-e^{i\omega}\big|^2
$$

$\rightarrow$ DOWN: FRma
$$
MA(1)\\
\Psi(B)=1+\theta B\\
f(\omega)=1+2\theta\cdot cos\theta+\theta^2=\big|\Psi(e^{i\omega})\big|^2
$$